---
title: "M4: Outliers and Missing Data"
author: "Luana Lima"
output: pdf_document
---


## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)

#New packages for M4
#install.packages("outliers")
library(outliers)
#install.packages("tidyverse")
library(tidyverse)

```

## Importing and preparing data

Today we will work with wind speed data. 

```{r}

#Importing time series data from text file#
wind_data <- read.csv(file="../Data/Wind_Speed_PortArthurTX_Processed.csv",header=TRUE,stringsAsFactors = TRUE)

#creating data object
wind_data$DATE <- ym(wind_data$DATE)

```


## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_wind_data <- ts(wind_data$AWND.1,frequency=12) 

```

## Initial plots

Common plots for outlier detection are histograms and boxplots. Histograms will help you understand the shape and spread of the data and to identify any potential outliers. And boxplots will give more information on the spread of the data. 

```{r}
#using package ggplot2 to make histograms

ggplot(wind_data, aes(x = DATE, y = AWND.1)) +
  geom_line() +
  ylab("Average Wind Speed") 

```
```{r}
#There is a gap between around 1985, some are around other places in the graph
```

## Using pre-built function to clean series

tsclean() from package 'forecast' identify and replace outliers and missing values in a time series using linear interpolation.

```{r}

clean_wind_data <- tsclean(ts_wind_data)  #object should be a time series #tsclean connect the gap

full_wind_data <- data_frame(Date=wind_data$DATE, Wind = wind_data$AWND.1, Wind_Clean = as.numeric(clean_wind_data))

ggplot(full_wind_data) +
  geom_line(aes(x = Date, y = Wind_Clean), color = "red") +
  geom_line(aes(x = Date, y = Wind), color = "black") +
  ylab("Average Wind Speed") 
  

```


## Decomposing the time series

The stats package has a function called decompose(). This function only take time series object. As the name says the decompose function will decompose your time series into three components: trend, seasonal and random. This is similar to what we did in the previous script, but in a more automated way. 

The random component is the time series without seasonal and trend component. Let's try to identify outliers by looking at the random component only.

Additional info on decompose()

1) You have two options: alternative and multiplicative. Multiplicative models exhibit a change in frequency over time.

2) The trend is not a straight line because it uses a moving average method to detect trend. 

3) The seasonal component of the time series is found by subtracting the trend component from the original data then grouping the results by month and averaging them. 

4) The random component, also referred to as the noise component, is composed of all the leftover signal which is not explained by the combination of the trend and seasonal component. 

```{r}
#Ex1: Using R decompose function decompose the wind data in seasonal, trend and random
decompose_wind_data <- decompose(clean_wind_data) # the default is additive, can change to multiplicative (diving the original data by the trend); multiplicative assume that seasonal ocmponent will depend on the level of series
plot(decompose_wind_data)
#not a linear model, it is a moving average model

#Ex2: Store the three component in separate vector
wind_trend <- decompose_wind_data$trend
wind_seasonal <- decompose_wind_data$seasonal
wind_random <- decompose_wind_data$random

#Ex3: Create one data frame with all three vector and the Date columns from wind_data
df_wind_data <- data.frame("date"= wind_data$DATE, wind_trend, wind_seasonal, wind_random)

#Ex4: Inspect the data frame using head() and tail()
head(df_wind_data,24)  #loss 6 for the trend (moving average), random : original series - seasonal - trend =NA #for seasonal we just finding the means, can use the means of first observation; use in between observation to calculate the mean (0.446 repeat at 13) 
tail(df_wind_data,10) #see the same 6 NA, don't have next 6 observation for moving average
# Discuss the results

```

```{r}
decompose_wind_data <- stl(clean_wind_data, s.window = "periodic") #only additive model 
plot(decompose_wind_data)

#the smaller the bar, the more is representing your time series, the trend component is not very pronoucing, the seasonal component  represent most of the variable of your series
# if the bar is high: have to zoom in to see my dataset 
```

