---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 6 - Due date 03/06/23"
author: "Qiuying Liao"
output: pdf_document
geometry: margin=2.54cm
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

## Questions 

This assignment has general questions about ARIMA Models. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
```


## Q1
Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models: 


* AR(2)

> Answer: The AR(2) shows a strong autocorrelation and a gradual decay of autocorrelation with increasing lag in the acf of the model. The pacf of AR(2) will identify the order of the AR model (p=2), which shows significant spikes in the first two lags and cutoff after the lag 2.


* MA(1)

> Answer: The acf of MA(1) will identify the order of the model (q=1), meaning that there is a significant spike at lag=1 and cutoff after lag 1. The pacf of MA(1) will decay exponentially. 



## Q2
Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$. Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use R to generate $n=100$ observations from each of these three models

```{r}
#ARMA(1,0)
ARMAmodel_1<- arima.sim(model=list(ar=0.6), n=100) #the AR coefficient is 0.6
ARMAmodel_1

#ARMA(0,1)
ARMAmodel_2<- arima.sim(model=list(ma=0.9), n=100) #the MA coefficient is 0.9
ARMAmodel_2

#ARMA(1,1)
ARMAmodel_3 <- arima.sim(model=list(ar=0.6, ma=0.9), n=100)
ARMAmodel_3
```

(a) Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r}
par(mfrow=c(1,3))
acf(ARMAmodel_1, lag.max = 40, main = "acf for model 1")
acf(ARMAmodel_2, lag.max = 40, main = "acf for model 2")
acf(ARMAmodel_3, lag.max = 40, main = "acf for model 3")
```


(b) Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r}
par(mfrow=c(1,3))
pacf(ARMAmodel_1, lag.max = 40, main = "pacf for model 1")
pacf(ARMAmodel_2, lag.max = 40, main = "pacf for model 2")
pacf(ARMAmodel_3, lag.max = 40, main = "pacf for model 3")
```

(c) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer: Model 1 is AR, model 2 is MA, model 3 is ARMA. Model 1 is AR model because ACF will decay exponentially with time, and we found that it is decaying with lags in the ACF of model 1. Model 2 is MA model because pacf has slow decay, while it is not obvious in the above graph. Model 3 is ARMA model because in the PACF initial values dependent on the AR followed by the decay due to the MA part. 

(d) Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: The AR model somewhat matches with the theoretical value. We set phi=0.6, meaning that the coefficient of AR should be 0.6. However, sometimes we can find that the pacf of lag 1 is 0.6 for model 1, but pacf of lag 1 is not 0.6 when we rerun the ARMA again. It didn't match with the theoretical value for MA model since we didn't find 0.9 at lag 1. For the ARMA model, ar coefficient at lag 1 in pacf didn't match with the theoretical value (0.6), but ma coefficient at lag 1 in acf matches with the thoretical value (0.9). 

(e) Increase number of observations to $n=1000$ and repeat parts (a)-(d).

```{r}
#ARMA(1,0)
ARMAmodel_1_new<- arima.sim(model=list(ar=0.6), n=1000) #the AR coefficient is 0.6
ARMAmodel_1_new

#ARMA(0,1)
ARMAmodel_2_new<- arima.sim(model=list(ma=0.9), n=1000) #the MA coefficient is 0.9
ARMAmodel_2_new

#ARMA(1,1)
ARMAmodel_3_new<- arima.sim(model=list(ar=0.6, ma=0.9), n=1000)
ARMAmodel_3_new

```

```{r}
par(mfrow=c(1,3))
acf(ARMAmodel_1_new, lag.max = 40, main = "acf for model 1")
acf(ARMAmodel_2_new, lag.max = 40, main = "acf for model 2")
acf(ARMAmodel_3_new, lag.max = 40, main = "acf for model 3")
```

```{r}
par(mfrow=c(1,3))
pacf(ARMAmodel_1_new, lag.max = 40, main = "pacf for model 1")
pacf(ARMAmodel_2_new, lag.max = 40, main = "pacf for model 2")
pacf(ARMAmodel_3_new, lag.max = 40, main = "pacf for model 3")
```

(c) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer: Model 1 is AR, model 2 is MA, model 3 is ARMA. Model 1 is AR model because ACF will decay exponentially with time, and we found that it is decaying with lags in the ACF of model 1. Model 2 is MA model because pacf has an obvious slow decay. Model 3 is ARMA model because in the PACF initial values dependent on the AR followed by the decay due to the MA part

(d) Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: The AR model matches with the theoretical value. We set phi=0.6, meaning that the coefficient of AR should be 0.6, we can find that the pacf of lag 1 is 0.6 for model 1. It didn't match with the theoretical value for MA model since we didn't find 0.9 at lag 1. For the ARMA model, ar coefficient at lag 1 in pacf didn't match with the theoretical value (0.6) and ma coefficient at lag 1 in acf didn't match with the thoretical value (0.9). 


## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$


(a) Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

> Answer: ARIMA(1,0,1)(1,1,0)


(b) Also from the equation what are the values of the parameters, i.e., model coefficients. 

> Answer: AR coefficient: 0.7 , MA coefficient: 0.1, SAR coefficient: -0.25

## Q4

Plot the ACF and PACF of a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using R. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot ACF and PACF for the simulated data. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r}
#install.packages("sarima")
library(sarima)

SARIMAmodel_4<- sim_sarima(model=list(ma=0.5,sar=0.8, nseasons=12), n=1000) 
SARIMAmodel_4
```

```{r}
par(mfrow=c(1,2))
acf(SARIMAmodel_4, lag.max = 40, main = "pacf for SARIMA model")
pacf(SARIMAmodel_4, lag.max = 40, main = "pacf for SARIMA model")
```
Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

> Answer: From the non-seasonal part of the acf plot, it cuts off which is a MA process, and q=1. From the non-seasonal part of the pacf plot, it also tails off which shows a MA process, and p=2. From the seasonal part of the acf plot, there are multiple positive spikes at seasonal lags (lagg 12, 24, 36), showing that it is a SAR process and P=1. From the seasonal part of pacf plot, there is a single spike which also reflects a SAR process and Q=0.
