---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 3 - Due date 02/10/23"
author: "Qiuying Liao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
library(readxl)
library(ggplot2)
```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)

```{r}
#data import
library(readxl)

#Importing data set without change the original file using read.xlsx
energy_data <- read_excel(path="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 

#Now let's extract the column names from row 11
read_col_names <- read_excel(path="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 

colnames(energy_data) <- read_col_names
head(energy_data)
```

```{r}
#select columns for data frame 
energy_data_df <- energy_data[,c("Month", "Total Biomass Energy Production","Total Renewable Energy Production", "Hydroelectric Power Consumption")]

head(energy_data_df)
```

```{r}
#create time series
ts_energy_data_df <- ts(energy_data_df[,2:4], start = c(1973,1), frequency =12)
```

```{r}
#acf and pacf 

col_names <- colnames(energy_data_df)
cc <- palette()

for(i in 1:3){
  par(mfrow = c(1,3), mar = c(4,4,4,4))
  plot(ts_energy_data_df[,i], type="l", col=cc[i+2], ylab = paste0(col_names[i], "[Trillion Btu]"), main=col_names[i+1])
  Acf(ts_energy_data_df[,i], lag.max = 40, main = paste0("Acf of ",col_names[i]), ylim=c(-1,1)) #or main= name[i+1]
   Pacf(ts_energy_data_df[,i], lag.max = 40, main = paste0("Pacf of ",col_names[i]))#pasting the character HP and index i
}
```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?




### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r}
#Fit a linear trend to TS of iHP
nobs <- nrow(energy_data_df) 
t <- 1:nobs

linear_trend <- lm(ts_energy_data_df[,1] ~ t)
summary(linear_trend)

linear_trend2 <- lm(ts_energy_data_df[,2] ~ t)
summary(linear_trend2)

linear_trend3 <- lm(ts_energy_data_df[,3] ~ t)
summary(linear_trend3)

```

### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r}
beta0 <- linear_trend$coefficients[1]
beta1 <- linear_trend$coefficients[2]

#First 
detrend_energy_data_df <- ts_energy_data_df[,1]-(beta0+beta1*t)

ggplot(energy_data_df, aes(x=Month, y=ts_energy_data_df[,1]))+
  geom_line(color="blue")+
  geom_line(aes(y=detrend_energy_data_df),color="green")


beta0 <- linear_trend2$coefficients[1]
beta1 <- linear_trend2$coefficients[2]

#Second
detrend_energy_data_df2 <- ts_energy_data_df[,2]-(beta0+beta1*t)

ggplot(energy_data_df, aes(x=Month, y=ts_energy_data_df[,2]))+
  geom_line(color="blue")+
  geom_line(aes(y=detrend_energy_data_df2),color="red")

#Third
detrend_energy_data_df3 <- ts_energy_data_df[,3]-(beta0+beta1*t)

ggplot(energy_data_df, aes(x=Month, y=ts_energy_data_df[,3]))+
  geom_line(color="blue")+
  geom_line(aes(y=detrend_energy_data_df3),color="orange")




```


### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?

```{r acf}
name <- colnames(energy_data_df)

par(mfrow=c(1,2))
Acf(detrend_energy_data_df,lag.max = 40,main=paste0("Acf of ",col_names[i-1]),ylaim=c(-1,1))
Pacf(detrend_energy_data_df,lag.max = 40,main=paste0("Pacf of ",col_names[i-1]),ylim=c(-1,1))

par(mfrow=c(1,2))
Acf(detrend_energy_data_df2,lag.max = 40,main=paste0("Acf of ",col_names[i]),ylaim=c(-1,1))
Pacf(detrend_energy_data_df2,lag.max = 40,main=paste0("Pacf of ",col_names[i]),ylim=c(-1,1))

par(mfrow=c(1,2))
Acf(detrend_energy_data_df3,lag.max = 40,main=paste0("Acf of ",col_names[i+1]),ylaim=c(-1,1))
Pacf(detrend_energy_data_df3,lag.max = 40,main=paste0("Pacf of ",col_names[i+1]),ylim=c(-1,1))



```

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

```{r}
#Use seasonal means model
dummies <- seasonaldummy(ts_energy_data_df[,1]) #only have 11, have to leave one out for comparison
# run with intersect, then you will have 11

seas_means_model <- lm(ts_energy_data_df[,(1)]~dummies) #lm function only take data frame  # we used the orignal data (we detrend from original data in this case)
summary(seas_means_model)

#second
#Use seasonal means model
dummies <- seasonaldummy(ts_energy_data_df[,2]) #only have 11, have to leave one out for comparison
# run with intersect, then you will have 11

seas_means_model2 <- lm(ts_energy_data_df[,(2)]~dummies) #lm function only take data frame  # we used the orignal data (we detrend from original data in this case)
summary(seas_means_model2)

#third
#Use seasonal means model
dummies <- seasonaldummy(ts_energy_data_df[,3]) #only have 11, have to leave one out for comparison
# run with intersect, then you will have 11

seas_means_model3 <- lm(ts_energy_data_df[,3]~dummies) #lm function only take data frame  # we used the orignal data (we detrend from original data in this case)
summary(seas_means_model3)


```

### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?
```{r}
beta0 <- seas_means_model$coefficients[1]
beta1 <- seas_means_model$coefficients[2:12]

inflow_seas_comp <- array(0,nobs)
for(i in 1:nobs){
  inflow_seas_comp[i] = beta0 + beta1 %*% dummies[i,] #dummy only takes time series
}

ggplot(inflow_data,aes(x=my_date, y=inflow_data[,iHP+1]))+
  geom_line(color="blue")+
  geom_line(aes(y=inflow_seas_comp), color="orange")

```



### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?


